---
title: "Use of IlluminaHumanMethylationEPICv2anno.20a1.hg38"
author: "Zuguang Gu (z.gu@dkfz.de)"
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use of IlluminaHumanMethylationEPICv2anno.20a1.hg38}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE, message = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.align = "center")
```

This package provides annotations for Illumina methylation EPIC array v2.0. The data is based on the file https://support.illumina.com/content/dam/illumina-support/documents/downloads/productfiles/methylationepic/MethylationEPIC%20v2%20Files.zip from 
  https://support.illumina.com/array/array_kits/infinium-methylationepic-beadchip-kit/downloads.html.

When using with the **minfi** package, you can manually set the "annotation" element by providing the surfix (removing the string "IlluminaHumanMethylationEPICv2anno."):

```{r, eval = FALSE}
RGset = read.metharray.exp(...)

# explained in the IlluminaHumanMethylationEPICv2manifest package
annotation(RGset)["array"] = "IlluminaHumanMethylationEPICv2"

annotation(RGset)["annotation"] = "20a1.hg38"
```

## Compare EPIC array v1 and v2 probes

Next we compared CpG annotation in the package **IlluminaHumanMethylationEPICv2anno.20a1.hg38** and the package **IlluminaHumanMethylationEPICanno.ilm10b4.hg19**.


```{r}
library(IlluminaHumanMethylationEPICv2anno.20a1.hg38)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)

cgi1 = IlluminaHumanMethylationEPICanno.ilm10b4.hg19::Islands.UCSC
cgi2 = IlluminaHumanMethylationEPICv2anno.20a1.hg38::Islands.UCSC
```

Following code shows the number of probes in each CpG feature. Note in **IlluminaHumanMethylationEPICanno.ilm10b4.hg19**, CGI shores and shelves are additionally classified as `N_Shore`/`S_Shore` and `N_Shelf`/`S_Shelf`, while in **IlluminaHumanMethylationEPICv2anno.20a1.hg38** not.

```{r}
t1 = table(gsub("(N|S)_", "", cgi1$Relation_to_Island))
t1
t2 = table(cgi2$Relation_to_Island)
t2

t2 - t1
(t2 - t1)/t1
```

## Session info

```{r, echo = FALSE}
sessionInfo()
```
